---
title: Practical tools for quick data visualization - Introduction
author: miquel [dot] anglada [at] crg [dot] eu
date: 2021-02-17

font-size: 11pt
geometry: margin=1.0in
urlcolor: blue

output:
    html_document:
      self_contained: true
      toc: true
      toc_depth: true
      number_sections: true
---

When starting a new project, designing new methods for data analysis or even preparing
a new publication, data visualization is essential to both explore new hypotheses and
to validate our findings.

This crash talk aims to outline 

## Useful packages
- **Data Wrangling**
  - [`here`](https://cran.r-project.org/web/packages/here/vignettes/here.html)
  - [`readr`](https://readr.tidyverse.org/)
  - [`dplyr`](https://dplyr.tidyverse.org/)
  - [`tidyr`](https://tidyr.tidyverse.org/)
  - [`reshape2`](https://www.rdocumentation.org/packages/reshape2/versions/1.4.4)
  - [`seqinr`](https://seqinr.r-forge.r-project.org/)
  - [`ape`](https://www.rdocumentation.org/packages/ape/versions/5.4-1)

- **Visualization**
  - [`ggplot2`](https://ggplot2.tidyverse.org/)
  - [`ggpubr`](https://rpkgs.datanovia.com/ggpubr/)
  - [`pheatmap`](https://www.rdocumentation.org/packages/pheatmap/versions/1.0.12)
  - [`ggtree`](https://bioconductor.org/packages/release/bioc/html/ggtree.html)
  
## Install packages  
```{r install packages, message=FALSE, warning=FALSE}
# install packages from CRAN
## wrangling
install.packages('here')
install.packages('readr')
install.packages('dplyr')
install.packages('tidyr')
install.packages('reshape2')
install.packages('seqinr')
install.packages('ape')

## visualization
install.packages('ggplot2')
install.packages('ggpubr') # requires libcurl4 in ubuntu
install.packages('pheatmap')

# install packages from Bioconductor
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install('ggtree')
```

## Basic plots
- [bible](https://cran.r-project.org/web/packages/ggpubr/ggpubr.pdf)

### Bar plot
```{r}
```
### Pie plot
```{r}
```
### Histogram
```{r}
```
### Box plot
```{r}
```
### Strip plot
```{r}
```
### Dot plot
```{r}
```
### Violin plot
```{r}
```
### Scatter plot
```{r}
```
### Heat map
```{r}
```
### Phylogenetic tree
[book](https://guangchuangyu.github.io/ggtree-book/chapter-ggtree.html)
```{r}
library(ggtree)
set.seed(2017-02-16)
tree <- rtree(50)
ggtree(tree)
ggtree(tree, layout="slanted")
```


## Add annotations
### Stats
```{r}
```
### Labels
```{r}
```

## Other
- `ggparagraph`
- `ggsummarytable`

## Combine plots into figures
```{r}
```

## Save your figures
```{r}
```

## References

## Session Info
```{r session info}
sessionInfo()
```